---
title: "Analysis of Activity Monitoring Data"
author: "John Slough II"
date: "14 Jan 2015"
output: html_document
---
**Report**

A histogram with density curve of the average number of steps taken per day is shown below.
```{r,echo=TRUE}
setwd("~/Desktop/Coursera/Reproducible Research/project1")

activity=read.csv("activity.csv",header=TRUE)
activity$date=as.Date(activity$date)

library(plyr)
byDay=ddply(activity,"date",summarize, sum=sum(steps))

meanDay=round(mean(byDay$sum,na.rm=TRUE))
medianDay=median(byDay$sum,na.rm=TRUE)
```
```{r}
library(ggplot2)

ggplot(byDay, aes(x=sum))+ 
        geom_histogram(aes(y=..density..),binwidth=1000,
        colour="black", fill="lightblue") +
        geom_density(alpha=.2, fill="#FF6666")+
        labs(title="Histogram and Density of Average Number of Steps per Day")+
        labs(x="average number of steps per day",y="density")

```

The mean and median number of steps per day are 10766 and `r medianDay` respectively.  

**Average Daily Activity Pattern**

The following plot is a time series of the 5-minute interval and the average number of steps taken, averaged across all days.  


```{r}
byInt=ddply(activity,"interval",summarize, avg=mean(steps,na.rm=TRUE))

ggplot(byInt, aes(x = interval, y = avg, group = 1))+
         geom_line(colour="purple")+ 
        labs(title="Time Series of Average Number of Steps per 5-minute Interval")+
        labs(x="5 minute interval",y="average number of steps")

which.max(byInt[,2])
maxInt=round(byInt[104,])
byIntSort=byInt[order(byInt[,2],decreasing=TRUE),]
```

The interval which contains the maximum average number of steps taken is interval `r maxInt[,1]` for a maximum average of `r maxInt[,2]` number of steps.  

**Imputing Missing Values**

```{r}
miss=sum(is.na(activity))
n=nrow(activity)
prop=round(sum(is.na(activity))/nrow(activity)*100,1)
```
The total number of missing values in the dataset is `r miss` which corresponds to `r prop`% of the data.  
To fill in the missing data I will replace the missing value with the mean of the corresponding 5-minute interval for that missing value.  This is done with the packages 'plyr' and 'Hmisc'.  

```{r,}
library(Hmisc)
# create new dataset with imputed values
activity.imputed = ddply(activity, "interval", mutate, 
                         imputed.steps = impute(steps, mean))

act.imp.order=activity.imputed[order(activity.imputed[,2],decreasing=FALSE),]

activity.imp=act.imp.order[,c(4,2,3)]
activity.imp$imputed.steps=as.integer(activity.imp$imputed.steps)
detach("package:Hmisc")
```

A histogram with density curve of the new dataset with imputed missing values is shown below.

```{r}
byDay.imp=ddply(activity.imp,"date",summarize, sum=sum(imputed.steps))

ggplot(byDay.imp, aes(x=sum)) +theme_set(theme_bw())+ 
        geom_histogram(aes(y=..density..),binwidth=1000,
        colour="black", fill="lightblue") +
        geom_density(alpha=.2, fill="#FF6666")+
        labs(title="Histogram and Density of Average Number of Steps per Day
             (with Imputed Missing Values)")+
        labs(x="average number of steps per day",y="density")

mean.imp=round(mean(byDay.imp$sum,na.rm=TRUE))
median.imp=median(byDay.imp$sum,na.rm=TRUE)

```

The mean and median number of steps per day from the new dataset are 10750 and `r median.imp` respectively.  These values do differ from the estimates from the original dataset by `r 10766-10750` and `r medianDay-median.imp` for the means and medians, respectively.  From the histogram it appears that replacing the missing values with the means of that 5-minute interval have compressed the average total number of steps per day more around 10000 steps.  

**Differences in activity patterns between weekdays and weekends**

 The following plot, created by making a new factor variable for weekend and weekdays was created to show different activity patterns between these two factors.  
 

```{r}
# create factor variable for Weekend vs. Weekday
activity.imp$wend = as.factor(ifelse(weekdays(activity.imp$date) %in% 
        c("Saturday","Sunday"), "Weekend", "Weekday")) 

activity.WE=subset(activity.imp,wend=="Weekend")
activity.WD=subset(activity.imp,wend=="Weekday")

byWE.imp=ddply(activity.WE,"interval",summarize, avg=mean(imputed.steps))
byWD.imp=ddply(activity.WD,"interval",summarize, avg=mean(imputed.steps))
library(gridExtra)

ggWE=ggplot(byWE.imp, aes(x = interval, y = avg, group = 1))+ylim(0,250)+
        geom_line(colour="blue")+ 
        labs(title="Average Number of Steps per 5-minute Intervals\nWeekends")+
        labs(x="5 minute interval",y="average number of steps")
ggWE=ggWE+theme(plot.margin=unit(c(0,1,0,1), "cm"))

ggWD=ggplot(byWD.imp, aes(x = interval, y = avg, group = 1))+ylim(0,250)+
        geom_line(colour="salmon")+ 
        labs(title="Average Number of Steps per 5-minute Intervals\nWeekdays")+
        labs(x="5 minute interval",y="average number of steps")
ggWD=ggWD+theme(plot.margin=unit(c(0,1,0,1), "cm"))

grid.arrange(ggWE, ggWD, nrow=2, ncol=1)

```


We can clearly see that there is a higher spike in number of steps for the weekdays however, it appears that the number of steps is higher throughout the day for weekends.


**R-code**

Available at: https://github.com/SloughJE/Cousera-Regression-Models
